---
title: "Spam Detection"
output:
  word_document: default
  html_notebook: default
author: Rohit Rajagopal
---

### The following script utilises the SMS Spam Detection dataset from UCI to build various predictive models to categorise whether a message is spam or not. 

```{r}
# Load required libraries
library(tidyverse)
library(rpart)
library(randomForest)
library(e1071)
library(gbm)
library(stringr)
library(tm)
library(qdap)
library(textstem)
library(ggplot2)
library(wordcloud)
library(caret)
library(ranger)
set.seed(10)
```

**Data Loading**

Read in CSV file and modify dataset as required.

```{r}
# Read in dataset
sms_data <- read_csv("spam.csv")

# Find dataset dimensions
cat("Shape of Spam Dataset:", paste(dim(sms_data), collapse = "x"), "\n")
head(sms_data)
```

There seem to be a few blank columns, let's check these don't contain any values at all.

```{r}
# Count non-blank rows in each column using !is.na()
non_na_counts <- colSums(sapply(sms_data, function(x) !is.na(x)))
print(non_na_counts)
```
All the columns contain non-NA values, therefore let's concatenate columns 2 to 5 to form the entire message.

```{r}
# Create combined column and drop the remaining ones
sms_data$v2 <- paste(sms_data$v2,
                       ifelse(sms_data$...3 != "" & !is.na(sms_data$...3), sms_data$...3, ""),
                       ifelse(sms_data$...4 != "" & !is.na(sms_data$...4), sms_data$...4, ""),
                       ifelse(sms_data$...5 != "" & !is.na(sms_data$...5), sms_data$...5, ""),
                       sep = " ")
sms_data <- sms_data[, -c(3, 4, 5)]

# Rename remaining columns for readability
colnames(sms_data) <- c("spam_label", "text")
```

**Data Visualisation**

Let's have a look at how the dataset is distributed.

```{r}
# Find number of spam and non-spam messages
counts <- table(sms_data$spam_label)
print(counts)
```
```{r}
# Check and convert text encoding UTF-8
sms_data$text <- iconv(sms_data$text, to = "UTF-8", sub = "byte") 

# Find number of characters in text
sms_data$char_lengths <- nchar(sms_data$text)

# Create a histogram of characters in each message
ggplot(sms_data, aes(x = char_lengths, fill = spam_label)) +
  geom_histogram(binwidth = 10, position = "identity", alpha = 0.5, color = "black") +
  labs(title = "Histogram of Character Lengths by Spam Status",
       x = "Character Lengths",
       y = "Frequency") +
  scale_fill_manual(values = c("lightblue", "salmon")) +
  theme_minimal()

# Get summary stats
print(tapply(sms_data$char_lengths, sms_data$spam_label, summary))
```

```{r}
# Find number of words in text
sms_data$word_lengths <- str_count(sms_data$text, "\\S+")

# Create a histogram of words in each message
ggplot(sms_data, aes(x = word_lengths, fill = spam_label)) +
  geom_histogram(binwidth = 5, position = "identity", alpha = 0.5, color = "black") +
  labs(title = "Histogram of Word Lengths by Spam Status",
       x = "Word Lengths",
       y = "Frequency") +
  scale_fill_manual(values = c("lightblue", "salmon")) + 
  theme_minimal()

# Get summary stats
print(tapply(sms_data$word_lengths, sms_data$spam_label, summary))
```

```{r}
# Find whether each message contains numbers and if so, how many
sms_data$has_numbers <- grepl("\\d", sms_data$text)
sms_data$number_count <- sapply(gregexpr("\\d+", sms_data$text), function(x) sum(x > 0))

# Split of spam and non-spam messages containing numbers
number_count_splits = table(sms_data$spam_label, sms_data$has_numbers)
cat("\n")
print(number_count_splits)

# Create a histogram of the occurrences of numbers in each message
ggplot(sms_data, aes(x = number_count, fill = spam_label)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.5, color = "black") +
  labs(title = "Histogram of Number by Spam Status",
       x = "Number Occurrences",
       y = "Frequency") +
  scale_fill_manual(values = c("lightblue", "salmon")) + 
  theme_minimal()

# Get summary stats
cat("\n")
print(tapply(sms_data$has_numbers, sms_data$spam_label, summary))
cat("\n")
print(tapply(sms_data$number_count, sms_data$spam_label, summary))
```


```{r}
# Find whether each message contains non-alphanumeric characters and if so, how many
sms_data$has_non_alphanum <- grepl("[^[:alnum:]]", sms_data$text)
sms_data$non_alphanum_count <- sapply(gregexpr("[^[:alnum:]]", sms_data$text), function(x) sum(x > 0))

# Split of spam and non-spam messages containing non-alphanumeric characters
non_alphanum_count_splits = table(sms_data$spam_label, sms_data$has_non_alphanum)
cat("\n")
print(non_alphanum_count_splits)

# Create a histogram of the occurrences of non-alphanumeric characters in each message
ggplot(sms_data, aes(x = non_alphanum_count, fill = spam_label)) +
  geom_histogram(binwidth = 5, position = "identity", alpha = 0.5, color = "black") +
  labs(title = "Histogram of Non-alphanumeric Character Occurrences by Spam Status",
       x = "Non-alphanumeric Character Occurrences",
       y = "Frequency") +
  scale_fill_manual(values = c("lightblue", "salmon")) + 
  theme_minimal()

# Get summary stats
cat("\n")
print(tapply(sms_data$has_non_alphanum, sms_data$spam_label, summary))
cat("\n")
print(tapply(sms_data$non_alphanum_count, sms_data$spam_label, summary))
```

**Data Preprocessing**

Clean and process text data using a standardised NLP approach (such as removing punctuation, stop words, and lemmatisation), such that the models can interpret each message correctly.

```{r}
# Create a corpus (collection of text documents)
corpus <- Corpus(VectorSource(sms_data$text))

# Convert to lowercase
corpus <- tm_map(corpus, content_transformer(tolower))

# Remove numbers and punctuation
corpus <- tm_map(corpus, removeNumbers)
corpus <- tm_map(corpus, removePunctuation)

# Remove stopwords
corpus <- tm_map(corpus, removeWords, stopwords("english"))

# Perform lemmatisation (reduce words to their canonical or dictionary form in English)
corpus <- tm_map(corpus, lemmatize_strings)

# Remove extra white-space
corpus <- tm_map(corpus, stripWhitespace)
```

```{r}
# View first 5 messages in corpus to see cleaned output
for (i in 1:5) {
  line <- corpus[[i]][["content"]]
  cat(line, "\n")
}

# Check whether any message is now empty
for (i in 1:nrow(sms_data)) {
     if (nchar(corpus[[i]][["content"]]) == 0) {
         print(i)
     }
 }
```

```{r}
# Create a Document-Term Matrix (DTM)
dtm <- DocumentTermMatrix(corpus)
dtm_matrix = as.matrix(dtm)
spam_matrix <- dtm_matrix[which(sms_data$spam_label == "spam"), ]
ham_matrix <- dtm_matrix[which(sms_data$spam_label == "ham"), ]

# Convert dtm to a matrix of word frequencies
word_freq <- colSums(dtm_matrix)
spam_freq <- colSums(spam_matrix)
ham_freq  <- colSums(ham_matrix)

# Generate word clouds for spam and non-spam messages
spam_messages <- head(spam_freq, 75)
par(mfrow = c(1, 1)) 
wordcloud(names(spam_messages), spam_messages, min.freq = 1, random.order = FALSE, colors = brewer.pal(8, "Dark2"))
title(main = "Word Cloud for Top 75 Words used in Spam Messages")

ham_messages <- head(ham_freq, 75)
par(mfrow = c(1, 1)) 
wordcloud(names(ham_messages), ham_messages, min.freq = 1, random.order = FALSE, colors = brewer.pal(8, "Dark2"))
title(main = "Word Cloud for Top 75 Words used in Non-spam Messages")
```

```{r}
# Let's remove features to simplify model creation and likely pose no significance
#word_freq <- colSums(dtm_matrix)
#£threshold <- quantile(word_freq, probs = 0.5)
#top_words <- names(word_freq[word_freq >= threshold])
#dtm <- dtm[, top_words] # Retain only top 50% of features

# Compute TF-IDF weights
dtm_tfidf <- weightTfIdf(dtm)

# Convert to matrix format
word_matrix <- as.matrix(dtm_tfidf)

# Find matrix dimensions
cat("Shape of Word Matrix:", paste(dim(word_matrix), collapse = "x"), "\n")
```

**Model Creation**

Build 5 different models to predict whether a message should be classified as spam or not, and then evaluate their performance metrics.

```{r}
# Convert spam_label column to binary labels for model interpretation
sms_data$spam_label <- ifelse(sms_data$spam_label == "spam", 1, 0)

# Split data into train and test sets using a 80/20 split
indices <- sample(nrow(word_matrix), round(0.8 * nrow(word_matrix)))
x_train <- as.data.frame(word_matrix[indices, ])
colnames(x_train) <- make.names(colnames(x_train))
x_test <- as.data.frame(word_matrix[-indices, ])
colnames(x_test) <- make.names(colnames(x_test))
y_train <- sms_data$spam_label[indices]
y_test <- sms_data$spam_label[-indices]
```

Define models.

```{r}
# Train Naive Bayes model
nb_model <- naiveBayes(x_train, y_train)

# Train Decision Tree model
tree_model <- rpart(y_train ~ ., data = x_train, method = "class", control = rpart.control(maxdepth = 7))

# Train Random Forest model
rf_model <- ranger(y_train ~ ., data = x_train, num.tree = 100, mtry = sqrt(ncol(x_train)))

# Train SVM model (Linear kernel)
svm_model <- svm(x_train, y_train, kernel = "linear")

# Train Gradient Boosting Machine (GBM) model
gbm_model <- gbm(y_train ~ ., data = x_train, distribution = "bernoulli", n.trees = 100, interaction.depth = 3)
```

Predict on test set and evaluate model performance.

```{r}
# Predictions on test set
nb_pred <- predict(nb_model, newdata = x_test)
tree_pred <- predict(tree_model, newdata = x_test)
rf_pred <- predict(rf_model, newdata = x_test)
svm_pred <- predict(svm_model, newdata = x_test)
gbm_pred <- predict(gbm_model, newdata = x_test)

# Confusion matrices for test set
confusionMatrix(nb_pred, y_test)
confusionMatrix(tree_pred, y_test)
confusionMatrix(rf_pred, y_test)
confusionMatrix(svm_pred, y_test)
confusionMatrix(gbm_pred, y_test)
```











